Drug sensitivity prediction with normal inverse Gaussian shrinkage informed by external data

### 2020-04-23 docs/manuscript.Rnw
commit before pull

	SM Section \ref{sm-sec:simulations} contains more 
	simulation results for Scenarios 1-4 for
	the NIG model and four alternative methods: frequentist (i) 
	lasso and (ii) ridge models, and EB version of the ridge model: (iii) a
	multivariate Bayesian extension of \texttt{xtune} 
	\cite[]{ni_bayesian_2019} named mxtune, as introduced in SM Section 
	\ref{sm-sec:mxtune}. Additionally, SM Section \ref{sm-sec:simulations} 
	contains a comparison of MCMC and VB posteriors in simulation Scenario 3.

	\subsection{Results}
<<simulation_gdsc>>=
# simulation 1
res <- read.table("results/simulations_gdsc_res1.2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

@@

# simulation 2
res <- read.table("results/simulations_gdsc_res2.2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

@@

# simulation 3
res <- read.table("results/simulations_gdsc_res3.2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

@@

tab5.2 <- tab5[, c(4, 13, 14, 6:8, 10:12)]
tabm5.2 <- tabm5[, c(4, 13, 14, 6:8, 10:12)]
tab5.2 <- t(sapply(1:nrow(tab5.2), function(r) {
  id <- which(tabm5.2[r, ]==min(tabm5.2[r, ], na.rm=TRUE))
  out <- tab5.2[r, ]
  out[id] <- paste0("\\textbf{", out[id], "}")
  out}))
rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
colnames(tab5.2) <- c("lasso", "lasso(25)", "lasso(100)", 
                      "NIG$_{\\text{f}+\\text{d}}+$DSS(lasso)",
                      "NIG$_{\\text{f}+\\text{d}}+$DSS(25)",
                      "NIG$_{\\text{f}+\\text{d}}+$DSS(100)",
                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(lasso)",
                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(25)",
                      "NIG$_{\\text{f}+\\text{d}}^-+$DSS(100)")
tab5.2 <- t(tab5.2)

@@

  tab1, align="r", digits=2,

@@

  tab2, align="r", digits=2,

@@

  tab3, align="r", digits=2,

@@

  tab4.1, align="r", digits=2,

@@

  tab4.2, align="r", digits=2,

@@

  Part of the differences in performance between
  NIG and ridge may be explained with the different levels of 
  sparsity in the solution. Although NIG does not automatically select features, 
  as opposed to the lasso,
  its $\bm{\beta}_d$ prior has larger kurtosis than the ridge prior (see SM
  Section \ref{sm-sec:prior}). The resulting heavy-tailedness as compared to 
  the ridge prior facilitates features selection. 
  Selection of features from a sparse prior
  may be achieved through the decoupling shrinkage and selection approach (DSS)
  introduced in \cite{hahn_decoupling_2015}. We have applied DSS in our 
  analyses, where we select either (approximately) the same number of features 
  as lasso, 25, or 50 features. Table 
  \ref{tab:analysis_gdsc_pmse2} compares the resulting mean PMSE values. The
  Table shows that NIG$_{\text{f}+\text{d}}$ plus DSS outperforms lasso,
  with the same number of features.

<<analysis_gdsc_pmse2>>=  
kableExtra::kable_styling(knitr::kable(
  tab5.2, align="r",
  caption="Mean (standard deviation) of cross-validated PMSE for GDSC data. Best performing model (per analysis) in bold.",
  format="latex", booktabs=TRUE, escape=FALSE), 
  latex_options=c("HOLD_position"))
@

>>>>

	SM Section \ref{sm-sec:simulations} contains more 
	simulation results for Scenarios 1-4 for
	the NIG model and the (i) frequentist 
	lasso and (ii) ridge models. Additionally, SM Section \ref{sm-sec:simulations} 
	contains a comparison of MCMC and VB posteriors in simulation Scenario 3.

	\subsection{Results}
<<simulation_gdsc>>=
# simulation 1
res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

@@

# simulation 2
res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

@@

# simulation 3
res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

@@

tab5.2 <- tab5[, c(4, 13:14, 10:12)]
tabm5.2 <- tabm5[, c(4, 13:14, 10:12)]
tab5.2 <- t(sapply(1:nrow(tab5.2), function(r) {
  id <- which(tabm5.2[r, ]==min(tabm5.2[r, ], na.rm=TRUE))
  out <- tab5.2[r, ]
  out[id] <- paste0("\\textbf{", out[id], "}")
  out}))
rownames(tab5.2) <- paste0("analysis ", 1:nrow(tab5.2))
colnames(tab5.2) <- c("lasso($\\sim$50)$^*$", "lasso(25)", "lasso(100)", 
                      "NIG$_{\\text{f}+\\text{d}}+$DSS($\\sim$50)$^*$",
                      "NIG$_{\\text{f}+\\text{d}}+$DSS(25)",
                      "NIG$_{\\text{f}+\\text{d}}+$DSS(100)")
tab5.2 <- t(tab5.2)
tab5.2 <- tab5.2[c(2, 1, 3, 5, 4, 6), ]

@@

  tab1, align="r", digits=3,

@@

  tab2, align="r", digits=3,

@@

  tab3, align="r", digits=3,

@@

  tab4.1, align="r", digits=3,

@@

  tab4.2, align="r", digits=3,

@@

  Part of the differences in performance between
  NIG and ridge may be explained with the different levels of 
  sparsity in the solution. Although NIG does not automatically select features, 
  as opposed to the lasso,
  its $\bm{\beta}_d$ prior has larger kurtosis than the ridge prior (see SM
  Section \ref{sm-sec:prior}). The resulting heavy-tailedness as compared to 
  the ridge prior facilitates features selection. 
  Selection of features from a sparse prior
  may be achieved through the decoupling shrinkage and selection approach (DSS)
  introduced in \cite{hahn_decoupling_2015}. We have applied DSS in our 
  analyses, where we select either (approximately) the same number of features 
  as lasso (about 50 in all analyses), 25, or 100 features. Table 
  \ref{tab:analysis_gdsc_pmse2} compares the resulting mean PMSE values. The
  Table shows that NIG$_{\text{f}+\text{d}}$ plus DSS outperforms lasso,
  for all three numbers of selected features, in all analyses.

<<analysis_gdsc_pmse2>>=  
kableExtra::add_footnote(kableExtra::kable_styling(knitr::kable(
  tab5.2, align="r",
  caption="Mean (standard deviation) of cross-validated PMSE for selection methods (number of selected features between parentheses) on GDSC data. Best performing model (per analysis) in bold.",
  format="latex", booktabs=TRUE, escape=FALSE), 
  latex_options=c("HOLD_position")), 
  label="cross validated lasso selects, on average, 42-56 features in all analyses (indicated with $\\sim$50).",
  notation="symbol", threeparttable=TRUE, escape=FALSE)
@

### 2020-04-23 docs/supplement.Rnw
commit before pull

	\section{Content overview}
	This document contains the Supplementary Material (SM) to the document 
	`Drug sensitivity prediction with the normal inverse Gaussian prior 
	informed by external data'. In the 
	following, this document is referred to as Main Document (MD).

	Section \ref{sec:prior} describes the moments of the normal
	inverse Gaussian (NIG) prior, Section \ref{sec:vbderivations} gives
	the variational Bayes (BV) derivations that lead to the estimating equations 
	in the MD, Section \ref{sec:elbo} gives the evidence lower bound (ELBO) that 
	is maximised during the VB iterations. In Section \ref{sec:computation} 
	efficient VB calculations are given, while Section 
	\ref{sec:ratiosmodifiedbessels} outlines a computationally stable method
	to calculate ratios of modified Bessel functions. Section \ref{sec:eb} shows
	that the empirical Bayes (EB) scale estimates $\bm{\lambda}$ are non-negative.
	The last Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
	to sample from the posterior corresponding to the NIG model.

@@

  \section{GDSC data}\label{sec:gdsc}
  the CPO for cell line $i$ and drug $d$ is calculated as 
  $$
  \text{CPO}_{id} = p(y_{id} | \bm{y}_{-id}) = 
  \[\int_{\bm{\beta}_d} \int_{\sigma_d^2} \mathcal{L}(\y_{id} | 
  \bm{\beta}_d, \sigma_d^2)^{-1} q(\bm{\beta}_d | \bm{y}_d) 
  q(\sigma^2_d | \bm{y}_d) d\bm{\beta}_d \sigma_d^2\]^{-1}
  $$ 
  and is the Bayesian version of the leave-one-out cross-validated likelihood.
  \cite{congdon_bayesian_2005} suggests as outlier cut-off 0.01

>>>>

	\section{Content overview}
	This document contains the Supplementary Material (SM) to the document 
	`Drug sensitivity prediction with normal inverse Gaussian shrinkage 
	informed by external data'. In the 
	following, this document is referred to as Main Document (MD).

	Section \ref{sec:prior} describes the moments of the normal
	inverse Gaussian (NIG) prior. Section \ref{sec:hyperparameters} describes
	the hyperparameter settings to generate Figure \ref{md-fig:dens_kappa} in the
	MD and Section \ref{sec:jeffreysprior} outlines the derivation to arrive at
	the Jeffreys prior for the error variance in equation (\ref{md-eq:sigmaprior})
	of the MD. Section \ref{sec:vbderivations} gives
	the variational Bayes (BV) derivations that lead to the estimating equations 
	in the MD, Section \ref{sec:elbo} gives the evidence lower bound (ELBO) that 
	is maximised during the VB iterations. In Section \ref{sec:computation} 
	efficient VB calculations are given, while Section 
	\ref{sec:ratiosmodifiedbessels} outlines a computationally stable method
	to calculate ratios of modified Bessel functions. Section \ref{sec:eb} shows
	that the empirical Bayes (EB) scale estimates $\bm{\lambda}$ are non-negative.
	Section \ref{sec:gibbssampler} introduces a Gibbs sampling scheme
	to sample from the posterior corresponding to the full Bayesian version of the
	NIG model. Finally, Sections \ref{sec:simulations} and \ref{sec:gdsc} contain
	additional results to the simulations and GDSC data analyses in Sections
	\ref{md-sec:simulations} and \ref{md-sec:gdsc} in the MD, respectively.

@@

  \section{GDSC data}\label{sec:gdsc}
<<analysis_gdsc>>=
sigma <- 0.83
cutoff <- 0.01
prop <- round(pnorm( - sigma*sqrt(-log(2) - log(pi) - 2*log(sigma) - 
                                    2*log(cutoff)), 0, sigma), 4)

load("results/analysis_gdsc_cpo1.Rdata")
prop1 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
load("results/analysis_gdsc_cpo2.Rdata")
prop2 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
load("results/analysis_gdsc_cpo3.Rdata")
prop3 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
load("results/analysis_gdsc_cpo4.Rdata")
prop4 <- round(mean(unlist(lcpo.nig2) < log(cutoff)), 4)
@
  the CPO for cell line $i$ and drug $d$ is
  \begin{align}\label{eq:cpo}
  \text{CPO}_{id} = p(y_{id} | \bm{y}_{-id}) = 
  \[\int_{\bm{\beta}_d} \int_{\sigma_d^2} \mathcal{L}(\y_{id} | 
  \bm{\beta}_d, \sigma_d^2)^{-1} q(\bm{\beta}_d | \bm{y}_d) 
  q(\sigma^2_d | \bm{y}_d) d\bm{\beta}_d \sigma_d^2\]^{-1}
  \end{align}
  and is the Bayesian version of the leave-one-out cross-validated likelihood.
  \cite{congdon_bayesian_2005} suggests as outlier cut-off 0.01.
  Figure \ref{fig:analysis_gdsc_cpo} displays these CPOs (calculated by 
  numerical integration of (\ref{eq:cpo})) for the four Analyses.
  For $y_{id} | y_{-id}$ following a Gaussian distribution with standard 
  deviation 0.83 
  (the average posterior mean of $\sigma^2_d$ from our analyses) this implies 
  an expected proportion of outliers of about \Sexpr{prop}. In our analyses we 
  find \Sexpr{prop1}, \Sexpr{prop2}, \Sexpr{prop3}, and \Sexpr{prop4}, for the
  four analyses respectively. So about double the amount of expected outliers
  under a Gaussian posterior predictive distribition. However, in reality,
  due to posterior uncertainty of $\bm{\beta}_d$ and $\sigma^2_d$, the
  predictive distribution is more heavy-tailed than the Gaussian and the 
  corresponding expected proportion of outliers higher than \Sexpr{prop}. 
  Unfortunately, calculation of this true expected proportion is infeasible 
  due to intractibility of the posterior predictive distribution.
<<analysis_gdsc_cpo, fig.cap="distribution of the CPO values for Analysis (a) 1, (b) 2, (c) 3, and (d) 4 in Section \\ref{md-sec:gdsc} of the MD, with outlier cutoff value 0.01 (red line).", out.width="100%">>= 
@
